\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-02-07 analysis/paper/casebase-manuscript.tex
update last table
Other minor changes

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given a
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \textbf{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
then describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis of
time-to-event data. We illustrate all these features through three
different case studies.\\
*SRB and MT contributed equally to this work.
}

@@

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. We fit
four models that differ in which functional form of time is used: 1)
excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear
function as seen in \eqref{eq:gomp}, 3) log function as seen in
\eqref{eq:weibull}, and 4) a smooth function using cubic B-splines. The
models are fit using \code{fitSmoothHazard} with the familiar formula
interface:

\begin{Schunk}
\begin{Sinput}
fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(i) {
  fitSmoothHazard(i, data = ERSPC, ratio = 100)
})
\end{Sinput}
\end{Schunk}

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. As such, we can directly
use the \code{summary} generic:

\begin{Schunk}
\begin{Sinput}
summary(fits[["splines"]]) 
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{Soutput}
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)
#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16
#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08
#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06
#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11
#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5785.3  on 54535  degrees of freedom
#> AIC: 5795
#> 
#> Number of Fisher Scoring iterations: 9
\end{Soutput}
\end{Schunk}

Next, the \code{absoluteRisk} function takes as input the
\code{fitSmoothHazard} object and returns a matrix where each column
corresponds to the covariate profiles specified in the \code{newdata}
argument, and each row corresponds to a specified time point:

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)

risks <- lapply(fits, function(i) {
  absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
})
\end{Sinput}
\end{Schunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
\pkg{casebase} on the Cox model CIF. The CIF estimates for the
exponential model in panel (1) overestimate the cumulative incidence
earlier on in time, and underestimate it later on. Based on our earlier
discussion of the population-time plot, this poor fit for the
exponential hazard was expected. We notice a better fit with increasing
complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
(panels 2--4). As noted above, the usual asymptotic results hold for
likelihood ratio tests built using case-base sampling models. Therefore,
we can easily test the null hypothesis that the exponential model is
just as good as the larger (in terms of number of parameters) spline
model:

@@

Previous re-analyses of these data suggested that the overall screening
attributed reduction in death due to prostate cancer of 20\% was an
underestimate \citep{hanley2010mortality}. The estimated 20\% (from a
proportional hazards model) did not account for the delay between
screening and the time the effect is expected to be observed. As a
result, the null effects in years 1--7 masked the substantial reductions
that began to appear from year 8 onwards. This motivates the use of a
time-dependent hazard ratio which can easily be fit with the
\pkg{casebase} package by including an interaction term with time in the
model:

@@

With this first case study, we explored how \pkg{casebase} allows us to
fit different parametric survival models with possible time-varying
effects, and how we can compare the fit of each model with tools from
GLMs.

@@

@@

In this third case study, we showed how case-base sampling can be used
in conjunction with penalized logistic regression to perform variable
selection in survival models.

@@

In this article, we presented the \textsf{R} package \pkg{casebase},
which provides functions for fitting smooth parametric hazards and
estimating CIFs using case-base sampling. Our package also provide
several functions to produce graphical summaries of the data and the
results. We outlined the theoretical underpinnings of the approach, we
provided details about our implementation, and we illustrated the merits
of the approach and the package through three case studies.

As a methodological framework, case-base sampling is very flexible. Some
of this flexibility has been explored before in the literature: for
example, Saarela and Hanley \citeyearpar{saarela2015case} used case-base
sampling to model a time-dependent exposure variable in a vaccine safety
study. As another example, Saarela and Arjas
\citeyearpar{saarela2015non} combined case-base sampling and a Bayesian
non-parametric framework to compute individualized risk assessments for
chronic diseases. In the case studies above, we further explored this
flexibility along two fronts. On the one hand, we showed how splines
could be used as part of the linear predictor to model the effect of
time on the hazard. This strategy yielded estimates of the survival
function that were qualitatively similar to semiparametric estimates
derived from Cox regression; however, case-base sampling led to
estimates of the survival function that \emph{vary smoothly in time} and
are thus easier to interpret. On the other hand, we also displayed the
flexibility of case-base sampling by showing how it could be combined
with penalized logistic regression to perform variable selection.
Furthermore, the second case study showed how case-base sampling can be
applied to competing risks settings. It should be noted that the three
case studies presented above only considered covariates that were fixed
at baseline. In the package vignette, we use the Stanford Heart
Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
to show how case-base sampling can also be used in the context of
time-dependent exposure. In this study, the exposure period was defined
as the week following vaccination. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://sahirbhatnagar.com/}%
%
\\\email{sahir.bhatnagar@mail.mcgill.ca}
}

>>>>

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given their
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \textbf{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can be used in more complex
settings: non-linear functions of time and non-proportional hazards,
competing risks, and variable selection. Our package also includes an
extensive array of visualization tools to complement the analysis of
time-to-event data. We illustrate all these features through three
different case studies.\\
* SRB and MT contributed equally to this work.
}

@@

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. We fit
four models that differ in which functional form of time is used: 1)
excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear
function (Eq. \eqref{eq:gomp}); 3) log function (Eq.
\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The
models are fit using \code{fitSmoothHazard} with the familiar formula
interface:

\begin{Schunk}
\begin{Sinput}
fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(form) {
  fitSmoothHazard(form, data = ERSPC, ratio = 100)
})
\end{Sinput}
\end{Schunk}

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we
leveraged the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

\begin{Schunk}
\begin{Sinput}
summary(fits[["splines"]]) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = form, data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.409  -0.172  -0.135  -0.084   3.808  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16 ***
#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08 ***
#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06 ***
#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11 ***
#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5785.3  on 54535  degrees of freedom
#> AIC: 5795
#> 
#> Number of Fisher Scoring iterations: 9
\end{Soutput}
\end{Schunk}

Next, the \code{absoluteRisk} function takes as input the
\code{singleEventCB} object and returns a matrix where each column
corresponds to the covariate profiles specified in the \code{newdata}
argument, and each row corresponds to time points specified by the
\code{time} argument:

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0, 14, by = 0.1)

risks <- lapply(fits, function(fit) {
  absoluteRisk(fit, time = new_time, newdata = new_data)
})
\end{Sinput}
\end{Schunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
\pkg{casebase} on the Cox model CIF. The CIF estimates for the
exponential model in panel (1) overestimate the cumulative incidence
earlier on in time, and underestimate it later on. Based on our earlier
discussion of the population-time plot, this poor fit for the
exponential hazard was expected. We notice a better fit with increasing
complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
(panels 2--4). As noted in Theoretical Details section, the usual
asymptotic results hold for likelihood ratio tests built using case-base
sampling models. Therefore, we can easily test the null hypothesis that
the exponential model is just as good as the larger (in terms of number
of parameters) spline model:

@@

Previous re-analyses of these data suggested that the overall screening
attributed reduction in death due to prostate cancer of 20\% was an
underestimate \citep{hanley2010mortality}. The estimated 20\% (from a
proportional hazards model) did not account for the delay between
screening and the time the effect is expected to be observed. As a
result, the null effects in years 1--7 masked the substantial reductions
that began to appear from year 8 onward. This motivates the use of a
time-dependent hazard ratio which can easily be fit with the
\pkg{casebase} package by including an interaction term with time in the
model:

@@

@@

\begin{Schunk}
\begin{Sinput}
model_cb <- fitSmoothHazard(
  Status ~ ftime + Sex + D + Phase + Source + Age,
  data = bmtcrr,
  ratio = 100,
  time = "ftime"
)
\end{Sinput}
\end{Schunk}

@@

@@

In this article, we presented the \textsf{R} package \pkg{casebase},
which provides functions for fitting smooth parametric hazards and
estimating CIFs using case-base sampling. Our package also provides
several functions to produce graphical summaries of the data and the
results. We outlined the theoretical underpinnings of the approach, we
provided details about our implementation, and we illustrated the merits
of the approach and the package through three case studies.

As a methodological framework, case-base sampling is very flexible. Some
of this flexibility has been explored before in the literature: for
example, Saarela and Hanley \citeyearpar{saarela2015case} used case-base
sampling to model a time-dependent exposure variable in a vaccine safety
study. As another example, Saarela and Arjas
\citeyearpar{saarela2015non} combined case-base sampling and a Bayesian
non-parametric framework to compute individualized risk assessments for
chronic diseases. In the case studies above, we further explored this
flexibility along two fronts. On the one hand, we showed how splines
could be used as part of the linear predictor to model the effect of
time on the hazard. This strategy yielded estimates of the survival
function that were qualitatively similar to semiparametric estimates
derived from Cox regression; however, case-base sampling led to
estimates of the survival function that \emph{vary smoothly in time} and
are thus easier to interpret. On the other hand, we also displayed the
flexibility of case-base sampling by showing how it could be combined
with penalized logistic regression to perform variable selection.
Furthermore, the second case study showed how case-base sampling can be
applied to competing risks settings. It should be noted that the three
case studies presented above only considered covariates that were fixed
at baseline. In one of the package vignettes, we use the Stanford Heart
Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
to show how case-base sampling can also be used in the context of
time-dependent exposure. In this study, the exposure period was defined
as the week following vaccination. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://sahirbhatnagar.com/}%
%
\\\email{sahir.bhatnagar@mcgill.ca}
}
